---
title: "R/Shiny Workshop"
author: "Daniel Krieg"
date: "10. Dezember 2019"
output: 
  rmdformats::material:
    highlight: kate
    css: style.css
  html_document:
    css: style.css
runtime: shiny
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)

source("solutions.R", local=TRUE)
```

# Introduction

## What is Shiny

A shiny app is a webbased user interface, which is created and connected to an active R session.
The session can react to user interaction and create/update content interactively.

### Pros

* No HTML knowledge required (while the basics might be helpful)
* Shiny is easy to learn and install
* Yields quick results for simple applications
* Can also be scaled for complex applications
* Use the full power of R for data processing and graphics


## Preparation



```{r echo=TRUE, eval=FALSE}

# Shiny
install.packages("shiny")

# Datatable: Nice interactive tables
install.packages("DT")

# Visualisations: 
install.packages("ggplot2")
install.packages("leaflet")

# Deploy to shinyapps
install.packages("rsconnect")

# Dashboard layout for Shiny
install.packages("shinydashboard")

# Themes for dashboards
install.packages("devtools")
devtools::install_github("nik01010/dashboardthemes")
```


# Basic structure

Create a file *app.R* containing 

* a variable `ui` describing the user interface (input and output elements)
* a function `server` to handle the user interaction (via `input`) and dynamically update the content (via `output`)
* a call to `shinyApp`

## Minimal example

```{r}
ui <- fluidPage(title = "Das ist der Browsertabtitel",
                h3("Erste leere Seite"),
                style=page_style)

server <- function(input, output, session) {}

shinyApp(ui = ui, server = server)
```


## Inputs

```{r}
library(shiny)

ui <- fluidPage(style=page_style,
      h3("Action Button"), 
      actionButton("action", label = "Action"),
      hr(),
      
      textInput("text", label = h3("Text input"), value = "Enter text..."),
      hr(),
      
      h3("Checkbox"),
      checkboxInput("checkbox", label = "Choice A", value = TRUE),
      hr(),
      
      checkboxGroupInput("checkGroup",
                         label = h3("Checkbox group"),
                         choices = list("Choice 1" = 1,"Choice 2" = 2,"Choice 3" = 3),
                         selected = 1),
      hr(),
      
      radioButtons("radio",
                   label = h3("Radio buttons"),
                   choices = list("Choice 1" = 1,"Choice 2" = 2,"Choice 3" = 3),
                   selected = 1),
      hr(),
      
      selectInput("select",
                  label = h3("Select box"),
                  choices = list("Choice 1" = 1,"Choice 2" = 2,"Choice 3" = 3),
                  selected = 1),
      hr(),
      
      numericInput("num",
                   label = h3("Numeric input"),
                   value = 1, min=0, max=10),
      hr(),
      
      sliderInput("slider1",label = h3("Slider"),min = 0,max = 100,value = 50),
      hr(), 
      
      sliderInput("slider2",label = h3("Slider Range"),min = 0,max = 100,value = c(40, 60))
    
  )

shinyApp(ui = ui, server = function(...){})
```


> Try it

We will create a page for displaying a monte carlo simulation of stock prices.

Start by constructing the inputs first:

```{r}


library(shiny)
library(dplyr)

shinyApp(ex1_ui, function(...){})

```

## Outputs

The content is generated in the *server* function by assign to elements of the `output` list.
Every `<obj>Output` in the *ui* has a corresponding `render<Obj>` function on the *server* side.

```{r}

library(shiny)
library(dplyr)
library(ggplot2)

ui <- fluidPage(style=page_style,
      h3("Plot output"), 
      plotOutput("plot"),
      hr(),
      
      h3("Table output"), 
      tableOutput("table"),
      hr(),
      
      h3("Data table output"), 
      dataTableOutput("datatable"),
      hr()
)

server <- function(input, output, session) {
  
  output$plot <- renderPlot({
    data <- mtcars %>% 
      mutate(cyl = as.factor(cyl))
    ggplot(data, aes(qsec, mpg, color=cyl)) + geom_point()
  })
  
  output$table <- renderTable(head(mtcars, 10))
  
  output$datatable <- renderDataTable(mtcars, options = list(pageLength = 5))
}
      
shinyApp(ui = ui, server = server)

```



> Try it

Now add the outputs and render the plots in the server function.
Hint: Start by generating paths from fixed parameters, like:

```{r}
mean_return <- 0.05
volatility <- 0.1
iterations <- 100

price_paths <- simulate_paths(iterations, mean_return, volatility)
  
```


```{r}


library(shiny)
library(dplyr)

shinyApp(ex1_ui, ex1_server_fixed)

```





## Layout

A good introduction is given in the [Layout guide](https://shiny.rstudio.com/articles/layout-guide.html). 

### Grid

The basic layout is grid based (`fluidPage()`) and constructed row by row (`fluidRow()`) with nested columns (`columns()`)


```{r}
library(shiny)

ui <- fluidPage(style=page_style,
  fluidRow(
    column(2, "just text"),
    column(4, wellPanel(h3("Well panel with header"), "and content")),
    column(6, div("A div", style="background-color:#aa2222"))
  )
)

shinyApp(ui=ui, server = function(...){})

```


### Tabs and navigation



```{r}

page_1 <- fluidPage(style=page_style,
  titlePanel("Hello Shiny!"),
  
  
)

page_2 <- fluidPage(style=page_style,
  tabsetPanel(
    tabPanel("Plot", wellPanel("Plot Content")), 
    tabPanel("Summary", wellPanel("Summary Content")), 
    tabPanel("Table", wellPanel("Table Content"))
  )
)

page_3 <- fluidPage(style=page_style,
  navlistPanel(
    "Output",
    tabPanel("Plot", wellPanel("Plot Content")), 
    "Data",
    tabPanel("Summary", wellPanel("Summary Content")), 
    tabPanel("Table", wellPanel("Table Content"))
  )
)



ui <- navbarPage("My Application",
                 tabPanel("Start", page_1),
                 tabPanel("Tabs", page_2),
                 tabPanel("NavLists", page_3)
)


shinyApp(ui=ui, server = function(...){})


```


> Try it

Reuse the code from the monte carlo example and modify it.
Put the inputs side by side (Hint: use `column`) and the outputs in tabs (Hint: use `tabsetPanel` and `tabPanel`).

```{r}


library(shiny)
library(dplyr)

shinyApp(ex2_ui, ex1_server_fixed)

```


# Reactive Programming

In order to create an interactive and efficient UI, the server needs to

* read the inputs (reactive sources) from `input`
* optionally calculate intermediate objects for further use (reactive conducters)
* be able to recognize if anything changed and which are the dependent objects (invalidation)
* and update only the relevant output (reactive endpoint) via `output`

To employ the input values from the UI within the server side, refer to them via the `input` list.
They are no Base R objects, but so-called reactive expressions. They know on which other reactive source they depend. You need to evaluate them within in reactive context (e.g. `renderText`).

## Reactive sources and endpoints 

**Hint:** Open the examples in a regular browser and press *Ctrl+F3* to open the reactlog.

```{r}
library(shiny) 

options(shiny.reactlog=TRUE) 

ui <- fluidPage(
  textInput("text_input", "Text"),
  verbatimTextOutput("text_output"))

server <- function(input, output, session) {
  
  # Not working:
  # output$text_output <- input$text_input
  
  output$text_output <- renderText({
    paste("my text:", input$text_input)
  })

}

shinyApp(ui, server)

```


## Reactive conducters

Conducters are reactive objects which depend on reactive input and have other reactive conducters and endpoints depending on them.

```{r}
library(shiny) 

options(shiny.reactlog=TRUE) 

ui <- fluidPage(
  textInput("text_input", "Text"),
  actionButton("update", "Update"),
  verbatimTextOutput("text_output"),
  verbatimTextOutput("text_output2"))

server <- function(input, output, session) {
  
  # Create a new reactive conducter
  doubled_text <- reactive({
    paste(input$text_input, input$text_input)
  })
  
  output$text_output2 <- renderText({
    # Not working:
    # doubled_text
    
    # Notice the brackets
    paste("my doubled text:", doubled_text())
  })
  
  ## Create a reactive conducter depending on another source
  
  text_delayed <- eventReactive(input$update, {input$text_input}  )
  
  output$text_output <- renderText({
    paste("my delayed text:", text_delayed())
  })
  
  
  
}

shinyApp(ui, server)

```


> Try it

Reuse the monte carlo example, but this time create a reactive conducter, which holds the simulated paths and depends on the inputs.

```{r}


library(shiny)
library(dplyr)

shinyApp(ex2_ui, ex1_server_reactive)


```


### Relevant reactive functions

Conducters:

* `reactive(expr, ...)`:
Creates a reactive conducter which invalidates on *expr*
* `eventReactive(eventExpr, valueExpr, ...)`:
Creates reactive expression with code in *valueExpr* that only invalidates when reactive values in *eventExpr* change.
* `isolate(expr)`:
Runs a code block. Returns a non-reactive copy of the results, i.e. no invalidation dependency.

Endpoints:

* `observe(expr, ...)`:
Runs code *expr* when reactive values change
* `observeEvent(eventExpr, handlerExpr)`: 
Runs code in *handlerExpr* when reactive values in *eventExpr* change. 

Delays:

* `throttle(expr, d)`: Delay invalidation such that it does not takes place more often than every *d* milliseconds
* `debounce(expr, d)`: Prevent multiple invalidations until no invalidation happend within a time frame of *d* milliseconds



> Try it

Plotting the paths is the expensive part if iterations are large. 
Add a button to refresh the paths if the inputs change.

*Hint*: Start with the button and toggle its state. Then use the button for updating the plot.

```{r}


library(shiny)
library(dplyr)

shinyApp(ex3_ui, ex3_server_reactive)


```


# Scalable and production-ready applications

To create robust applications it is important that your code is clean and well organizied and doesn't break on any changes.

Therefore, as applications get bigger and more complex, you need 

* modular code
* automated testing

Especially for a reactive user interface application like Shiny, testing can be very time consuming.

## Shinytest

The *shinytest* package offers an automated testing based on a scripted interaction with the application.

In addition to the package, you need the external application *phantomJS* installed and in your path.

```{r echo=TRUE, eval=FALSE}
install.packages("shinytest")

shinytest::installDependencies()
```


The workflow is as follows

1. Record a test:  
  Interact with the application and define relevant snapshots to take
2. Run the test the first time:  
  shinytest applies your recorded test script to store the expected results (state of the application data, screenshots)
3. Re-Run the test:  
  shinytest applies your recorded test script and compares the results with expectations
  


# Ressources

- [RStudio Layout guide](https://shiny.rstudio.com/articles/layout-guide.html)
- [RStudio Shiny Cheat Sheet](https://shiny.rstudio.com/articles/cheatsheet.html)




# Backup



```{r}

ui <- navbarPage("My Application",
                 tabPanel("Page 1", uiOutput("page")),
                 tabPanel("Page 2", 
                          fluidPage(
                          conditionalPanel(
                            condition = "input.next > input.back",
                            actionButton("back", "Back")),
                          actionButton("next", "Next"),
                          conditionalPanel(
                            condition = "input.next - input.back == 1",
                            h3("Step 1")
                          ),
                          conditionalPanel(
                            condition = "input.next - input.back == 2",
                            h3("Step 2")
                          )))
)

server <- function(input, output, session) {
  
  output$page <- renderUI({
     tagList(
      sliderInput("n", "N", 1, 1000, 500),
      textInput("label", "Label")
    )
  })
  
  
}

shinyApp(ui=ui, server = server)

```



